\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Updates from Overleaf
Tried to clarify reversions

The properties of the mutations inferred in the Wide and Long
ARGs are summarised in Figure~\ref{fig:mutational_spectra}B. In both
cases we have a large number of mutations, and a majority of these
(Wide ARG: 62.55\%, Long ARG: 72.23\%) are private to a single
sample, i.e.\ on terminal branches.
Although the average number
of mutations per sample is small
(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}),
the average number per site
is large in both ARGs (Wide ARG: 41.23; Long ARG: 36.10; Table~\ref{tab:args}).
However, a lower median count (Wide ARG: 14; Long ARG: 13),
suggests that these high mutation counts are
partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
mutations in the Wide ARG; Figure~\ref{fig:breakpoint-distribution}), which
may be artefactual.
A significant issue with the current version of \texttt{sc2ts} is an excess
of inferred reversion mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions). \sally{[What confuses me is that I would think that none of these reversions would remain after the reversion push step of Figure 1D.  Is this because ``immediately ancestral'' is incorrect or because the problem is that some of these immediate reversions are created deeper in the tree than the newly added samples (i.e., uncorrected by the reversion push). If the former, maybe say ``reverting the state change of a recent ancestral mutation'', if the latter, maybe say  ``(see Section~\ref{sec:args} for precise definitions, note that the reversion push algorithm of Figure 1D only eliminates reversions in the newly added portion of the tree and not reversions around earlier nodes created by this algorithm)''.]}
The majority of these mutations are likely to be
artefactual, and a key goal of future development is to improve
our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
address this issue.

@@

Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
(nucleotides coded as `unknown' or masked out
as described in Section~\ref{sec:data_preprocessing})
and it is important to account for this missinging in a systematic way.
In \texttt{sc2ts} we automatically impute missing data as samples are added
into the ARG, using the LS model.
To do this, we assume that sites with missing nucleotides are uninformative
to the path probability, by setting the emission probability from any
state $(A,C,G,U,-)$ to `unknown' equal to 1 (though we could have chosen any constant).
As a result, emissions of unknown nucleotides will not contribute to
differences in path probabilities.
Once the most likely copying path is determined, we then attach the
sample to the ARG (see Figure~\ref{fig:overview_sc2ts} and following subsections).
For each newly attached sample we encode its nucleotide sequence by
recording mutations where it differs from the nucleotide sequence
of its parental node, but (importantly) ignoring sites with missing data
in the new sample.
Thus, once the newly added sequence has been attached to the ARG
any missing data is imputed from its parental node. There is therefore
no missing data in the ARG; all missing bases are ``hard called''
at attachment time.
Note that this approach is equivalent to
using state-of-the-art imputation
methods~\citep[e.g.][]{Browning2018-nk,Delaneau2019-wl} with a reference
panel consisting of all sequences in the ARG,
since these methods are also based on the LS model.
% Could say more here, but let's leave it.
Evaluating the accuracy of missing data imputation
using \texttt{sc2ts} is an important facet of future work.

>>>>

The properties of the mutations inferred in the Wide and Long
ARGs are summarised in Figure~\ref{fig:mutational_spectra}B. In both
cases we have a large number of mutations, and a majority of these
(Wide ARG: 62.55\%, Long ARG: 72.23\%) are private to a single
sample, i.e.\ on terminal branches.
Although the average number
of mutations per sample is small
(Wide ARG: 0.77, Long ARG: 1.38; Table~\ref{tab:args}),
the average number per site
is large in both ARGs (Wide ARG: 41.23; Long ARG: 36.10; Table~\ref{tab:args}).
However, a lower median count (Wide ARG: 14; Long ARG: 13),
suggests that these high mutation counts are
partly driven by some hypermutable sites (e.g., site 28,271 has over 7,000
mutations in the Wide ARG; Figure~\ref{fig:breakpoint-distribution}), which
may be artefactual.

The current version of \texttt{sc2ts} infers a
large number of ``reversion'' mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation (see Section~\ref{sec:args} for precise definitions).
These are symptomatic of both data quality issues
such as ``time travellers'' (Section~\ref{sec:filtering_time_travellers})
and problematic sites (e.g., 28,271 as discussed in
Section~\ref{sec:breakpoint_intervals}; see also
Figure~\ref{fig:pango_XB_gisaid_graph} for an example of multiple
reversions at this site), as well as indicating opportunities
for improvement in tree building hueristics
(Section~\ref{sec:parsimony-heuristics}).
For example, the current ``reversion push'' operation
only eliminates reversions in the newly added portion of the tree
and not reversions around earlier nodes created by this algorithm.

@@

Genetic data for SARS-CoV-2 contains substantial amounts of missingness,
(nucleotides coded as `unknown' or masked out
as described in Section~\ref{sec:data_preprocessing})
and it is important to account for this missingness in a systematic way.
In \texttt{sc2ts} we automatically impute missing data as samples are added
into the ARG, using the LS model.
To do this, we assume that sites with missing nucleotides are uninformative
to the path probability, by setting the emission probability from any
state $(A,C,G,U,-)$ to `unknown' equal to 1 (though we could have chosen any constant).
As a result, emissions of unknown nucleotides will not contribute to
differences in path probabilities.
Once the most likely copying path is determined, we then attach the
sample to the ARG (see Figure~\ref{fig:overview_sc2ts} and following subsections).
For each newly attached sample we encode its nucleotide sequence by
recording mutations where it differs from the nucleotide sequence
of its parental node, but (importantly) ignoring sites with missing data
in the new sample.
Thus, once the newly added sequence has been attached to the ARG
any missing data is imputed from its parental node. There is therefore
no missing data in the ARG; all missing bases are ``hard called''
at attachment time.
Note that this approach is equivalent to
using state-of-the-art imputation
methods~\citep[e.g.][]{Browning2018-nk,Delaneau2019-wl} with a reference
panel consisting of all sequences in the ARG,
since these methods are also based on the LS model.
% Could say more here, but let's leave it.
Evaluating the accuracy of missing data imputation
using \texttt{sc2ts} is an important facet of future work.
